\documentclass[acmsmall,screen,review,anonymous]{acmart}

\usepackage{cleveref}
\usepackage{cancel}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{todonotes}

% ── Cambria listing style ──────────────────────────────────────
\definecolor{kwcolor}{RGB}{0,0,160}
\definecolor{commentcolor}{RGB}{100,100,100}
\definecolor{strcolor}{RGB}{160,32,32}
\definecolor{paramcolor}{RGB}{160,80,0}

\lstdefinelanguage{cambria}{
  morekeywords={with,handle,handler,return,do,in,fun,rec,if,then,else,case,of,inl,inr,effect,finally},
  sensitive=true,
  morecomment=[l]{--},
  morestring=[b]",
  literate=
    {->}{$\to$}2
    {<-}{$\leftarrow$}2
    {~>}{$\leadsto$}2
    {=>}{$\Rightarrow$}2
    {++}{$\mathbin{+\!\!+}$}2
    {\$p}{{\textcolor{paramcolor}{\$p}}}2
    {\$loc}{{\textcolor{paramcolor}{\$loc}}}4
    {\$tid}{{\textcolor{paramcolor}{\$tid}}}4
    {\$node}{{\textcolor{paramcolor}{\$node}}}5
    {\$urn}{{\textcolor{paramcolor}{\$urn}}}4,
}

\lstset{
  language=cambria,
  basicstyle=\small\ttfamily,
  keywordstyle=\color{kwcolor}\bfseries,
  commentstyle=\color{commentcolor}\itshape,
  stringstyle=\color{strcolor},
  columns=flexible,
  keepspaces=true,
  xleftmargin=1em,
  escapeinside={(*}{*)},
  numbers=none,
  aboveskip=0.5em,
  belowskip=0.5em,
}

% ── Macros ──────────────────────────────────────────────────────
\newcommand{\cambria}{\textsc{Cambria}}
\newcommand{\eff}{\textsc{Eff}}

\newcommand{\void}{\mathsf{Void}}
\newcommand{\unit}{\mathsf{Unit}}
\newcommand{\tint}{\mathsf{Int}}
\newcommand{\tbool}{\mathsf{Bool}}
\newcommand{\tdouble}{\mathsf{Double}}
\newcommand{\tstring}{\mathsf{Str}}
\newcommand{\tunique}{\mathsf{Unique}}
\newcommand{\tlist}[1]{\mathsf{List}(#1)}

\newcommand{\tfun}[2]{#1 \to #2}
\newcommand{\tpair}[2]{#1 \times #2}
\newcommand{\teither}[2]{#1 + #2}
\newcommand{\thandler}[2]{#1 \Rightarrow #2}
\newcommand{\thandlerp}[3]{#1 \xRightarrow{#2} #3}

\newcommand{\comp}[2]{#1\;!\;#2}
\newcommand{\arity}[2]{#1 \to #2}

\newcommand{\op}{\mathsf{op}}
\newcommand{\ops}{\mathsf{ops}}
\newcommand{\ret}{\mathsf{return}\;}
\newcommand{\ddo}{\mathsf{do}\;}
\newcommand{\iin}{\;\mathsf{in}\;}
\newcommand{\iif}{\mathsf{if}\;}
\newcommand{\then}{\;\mathsf{then}\;}
\newcommand{\eelse}{\;\mathsf{else}\;}
\newcommand{\ccase}{\mathsf{case}\;}
\newcommand{\oof}{\;\mathsf{of}\;}
\newcommand{\inmark}[1]{\mathsf{in{#1}}\;}
\newcommand{\fun}{\mathsf{fun}\;}
\newcommand{\rec}{\mathsf{rec}\;}
\newcommand{\recmark}{\mathsf{rec}\;}
\newcommand{\hhandler}{\mathsf{handler}}
\newcommand{\with}{\mathsf{with}\;}
\newcommand{\handle}{\;\mathsf{handle}\;}
\newcommand{\effect}{\mathsf{effect}\;}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\cod}{\mathsf{cod}}
\newcommand{\ttrue}{\mathsf{true}}
\newcommand{\ffalse}{\mathsf{false}}
\newcommand{\pack}{\mathsf{pack}}
\newcommand{\unpack}{\mathsf{unpack}\;}
\newcommand{\as}{\;\mathsf{as}\;}

\newcommand{\G}{\Gamma}
\newcommand{\Sig}{\Sigma}
\newcommand{\ftv}{\mathsf{ftv}}
\newcommand{\fpv}{\mathsf{fpv}}
\newcommand{\rel}{\mathsf{rel}}

\newcommand{\tj}[3]{#1; #2 \vdash #3}
\newcommand{\vj}[5]{#1; #2; #3 \vdash #4 : #5}
\newcommand{\cj}[6]{#1; #2; #3 \vdash #4 : \comp{#5}{#6}}
\newcommand{\cjp}[5]{#1; #2; #3 \vdash #4 : #5}
\newcommand{\eenv}{\bullet}

% Step-indexed LR macros
\newcommand{\Rel}{\mathbf{Rel}}
\newcommand{\VV}{\mathcal{V}}
\newcommand{\SSS}{\mathcal{S}}
\newcommand{\CC}{\mathcal{C}}
\newcommand{\GG}{\mathcal{E}}

\newcommand{\ctx}{\mathcal{C}}

\setcopyright{none}
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL}
\acmArticle{1}
\acmYear{2027}
\acmMonth{1}
\acmDOI{}

\begin{document}

\title{Cambria: A Language for Parametrized Effects and Handlers}

\author{Anonymous}
\affiliation{\institution{Anonymous}}

\begin{abstract}
Parametrized algebraic theories are a generalization of algebraic theories where operations may bind abstract parameter types.
Models of these theories are essential for modelling computational effects with dynamically created resources, such as local state,
quantum computation, and dynamically threaded concurrency.
Algebraic effects and handlers is a programming language paradigm that separated the syntax from the semantics of computational effects,
allowing the programmer full control over their effect implementations.

We present \cambria{}, a langauage that extends the algebraic effects and handlers framework into the parametrized setting.
Parameters appear as abstract types introduced in the body of a handled computation and instantiated by the handler.
Parameters are first-class in the type system but erased at runtime, requiring no coersions or type-directed reduction.

We formally prove the abstraction guaruntee via a step-indexed logical relation to establish parametricity.
The handled computation cannot inspect the concrete instantiation.
We also establish type safety, and a minimal annotation result for completeness of the type inference algorithm.
We demonstrate \cambria{}'s effectiveness with a working implementation, and provide examples including local state,
probabilistic programming, and concurrent thread management. The latter is a parametrized effect that does not reduce to
effect instances, and thus this is the first language which can express such an effect in the handler framework.
\end{abstract}

\maketitle

%% ═══════════════════════════════════════════════════════════════
\section{Introduction}\label{sec:intro}
%% ═══════════════════════════════════════════════════════════════

Algebraic effects and handlers~\cite{plotkin_handling_2009} provide a principled approach to programming with side effects. A computation performs abstract \emph{operations} (reading state, spawning threads, flipping coins), and a \emph{handler} gives these operations meaning by intercepting them and supplying an implementation. This separation of interface from implementation has made algebraic effects an increasingly popular foundation for effect systems in both research and practice~\cite{bauer_programming_2015, hillerström_continuation_2017, leijen_koka_2014}.

Many practically important effects are \emph{parameterized}: their operation signatures mention a type that varies across instances. Local mutable state is the canonical example --- a state handler manages memory cells whose \emph{address type} is an implementation detail. The handler knows addresses are integers (or unique tokens), but the client code should not: it should treat addresses as opaque identifiers, using only the operations the handler provides. This abstraction is essential for safety and modularity --- a client that forges addresses or inspects their representation can violate the handler's invariants.

\paragraph{The problem of instances.}
In unparameterized effect systems, the state effect has a single, global signature $\mathsf{get} : \unit \to \tint$, $\mathsf{set} : \tint \to \unit$. This is inadequate when a program creates \emph{multiple} memory cells, each with its own identity. The type of a cell's address --- the \emph{parameter} of the effect --- must be abstract from the client's perspective but concrete from the handler's.

Prior work addresses this ``problem of instances'' through existential types~\cite{biernacki2018handle} or explicit effect instance mechanisms. These approaches have two limitations. First, they impose a runtime cost: existential types require pack/unpack coercions, and instance-based systems need type-directed reduction rules. Second, they have an expressiveness ceiling: some parameterized effects --- notably concurrent thread management~\cite{kammar2025equational} --- cannot be reduced to simple effect instances, and thus fall outside their scope.

\paragraph{Parameter variables.}
\cambria{} takes a different approach. We extend the type system with \emph{parameter variables} $\psi$, written \lstinline{$p} in concrete syntax. A parameter variable is introduced in the body of a handled computation (via \lstinline{effect}) and instantiated by the handler:

\begin{lstlisting}
with handler { $cell -> Unique, get a k -> ..., set x k -> ..., ref x k -> ... }
handle (
  effect get : $cell ~> Int.
  effect set : $cell * Int ~> Unit.
  effect ref : Int ~> $cell.
  -- here $cell is abstract: the program cannot inspect it
  do a <- !ref 42 in !get a
)
\end{lstlisting}

Inside the body, \lstinline{$cell} is an abstract type. The program cannot inspect values of type \lstinline{$cell} --- it can only manipulate them through the declared operations. The handler instantiates \lstinline{$cell -> Unique}, revealing the representation to its own implementation but not to the client. This achieves the same data abstraction as existential types, but through a purely static mechanism: parameter variables are erased at runtime, requiring no coercions or type-directed reduction.

\paragraph{Contributions.} We make the following contributions:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Language design.} We present \cambria{}, a core calculus combining algebraic effect handlers, parameter variables, let-polymorphism, and general recursion (\Cref{sec:overview}--\Cref{sec:calculus}). Parameter variables provide type abstraction orthogonally to the effect handling mechanism, leading to a clean separation of concerns.

  \item \textbf{Parametricity.} We prove a parametricity theorem via a step-indexed logical relation (\Cref{sec:parametricity}). This is, to our knowledge, the first logical relation combining algebraic effect handlers with general recursion, universal polymorphism, and step-indexing. The proof is modular: each type constructor contributes an independent clause to the value relation, and each typing rule contributes an independent case to the fundamental theorem.

  \item \textbf{Type inference.} We establish a minimal annotation completeness result (\Cref{sec:inference}): only operations whose signatures mention instantiated parameters require type annotations. All other types, including polymorphic ones, are inferred.

  \item \textbf{Expressiveness.} We demonstrate that \cambria{} can express parameterized effects that prior work cannot, including concurrent thread management from recent work on dynamic threads~\cite{kammar2025equational}. We provide an implementation with runnable examples (\Cref{sec:overview}).
\end{enumerate}

%% ═══════════════════════════════════════════════════════════════
\section{Overview}\label{sec:overview}
%% ═══════════════════════════════════════════════════════════════

We introduce \cambria{} through a series of examples that exercise its key features: parameter variables, let-polymorphism, general recursion, and their interactions. Each example is executable in our implementation.

\subsection{Local State and Fresh Name Generation} \label{sec:local-state}

The following program declares the three operations involved in the local state effect and runs a program allocating and updating two memory cells:

\begin{lstlisting}
effect get : $loc ~> Int.
effect set : $loc * Int ~> Unit.
effect ref : Int ~> $loc.
do a <- !ref 2 in
do b <- !ref 3 in (
  !set (a, !get b); !get a
)
\end{lstlisting}

One possible method of handling this program is to instantiate the memory cells as maps from integers to the value.
Here, \lstinline|empty| is the empty map, and \lstinline|insert ((a, x), s)| inserts value \lstinline|x| at key \lstinline|a| in map \lstinline|s|.

\begin{lstlisting}
local_state_handler = handler {
  $loc -> Int,
  return x  -> return (fun _ -> return x),
  get a k -> return (fun s -> k (lookup (a, s)) s),
  set x k -> return (fun s -> k () (insert (x, s))),
  ref x k -> do a <- !fresh () in return (fun s -> k a (insert ((a, x), s))),
  finally s -> s empty
}
\end{lstlisting}

A fresh identifier for each memory location can be acquired by parameter passing an integer around much like a global state handler.
This shows that the parametrized handlers in \cambria{} integrate seamlessly with classic handlers, which need no updating to their syntax.

\begin{lstlisting}
fresh_handler = handler {
  $name -> Int,
  return x -> return (fun _ -> x),
  fresh _ k -> return (fun n -> k n (n+1)),
  finally f -> f 0
}
\end{lstlisting}

However, as we will see in the later sections, the effects and parameter instances may be instantiated in any way the programmer pleases.
\cambria{} guaruntees that the body cannot inspect the parameters of type \lstinline|$loc|.
It cannot compare them for equality, perform arithmetic, or forge new locations.
Even if the parameter \lstinline|$loc| is handled to in integer type,
the following code will not type-check because \lstinline|a| remains abstract until handled.

\begin{lstlisting}
effect get : $loc ~> Int.
effect set : $loc * Int ~> Unit.
effect ref : Int ~> $loc.
do a <- !ref 2 in !set (a+1, 3)
\end{lstlisting}

\subsection{Polymorphism Meets Abstraction}\label{sec:poly-refs}

Although the parameters in \cambria{} cannot be inspected, they still integrate seamlessly with let-polymorphism.
The following example illustrates a powerful pattern of mapping over a list of values, instantiating a memory location for each one,
and returning the list of locations.

\begin{lstlisting}
do map <- return (rec map (f, xs) ->
  case uncons xs of {
    inl _        -> return [],
    inr (x, xs') -> f x :: map f xs'
  }) in
with local_state_handler handle (
  effect get : $p ~> Int.
  effect ref : Int ~> $p.
  do refs <- map ((fun n -> !ref n), (10 :: 20 :: 30 :: [])) in
  do bools <- map ((fun r -> !get r == 20), refs) in
  return bools
)
\end{lstlisting}

The function \lstinline{map} has type $\forall\alpha_1\alpha_2.(\alpha_1 \to \comp{\alpha_2}{\Sig}) \times \tlist{\alpha_1} \to \comp{\tlist{\alpha_2}}{\Sig}$.
Even though the programmer cannot determine what the parameter \lstinline|loc| will be instantiated to, let-polymorphism is universally quantifying, and so
still capture the parameter types.

This combination of recursion, polymorphism, and parameter abstraction is not available in any prior handler calculus.
Biernacki et al.~\cite{biernacki2018handle} have no recursion or universal types.
The $\lambda^{\mathsf{HEL}}$ line of work~\cite{} adds instances but requires runtime coercions and lacks parametricity guarantees.

\subsection{Representation Independence: Two Implementations of Urns}\label{sec:polya}

The key idea behind parametrized handlers is that they provide a means for different implementations of the same effect interface,
and parametricity guarantees that the client code behaves the identically with both.

Polya's urn is a probabilistic process where an urn contains a red ball and a blue ball.
After each sample from the urn, an additional ball of the sampled colour is places back in the urn.
That is, the urn gets biased towards the previous samples.

Consider a probabilistic program that samples from a Polya urn ten times and accumulates the number of draws of red balls.
\begin{lstlisting}
effect newurn : Unit ~> $urn.
effect sample : $urn ~> Bool.
do urn <- !newurn () in
countRedDraws 10 (fun _ -> !sample urn) 
\end{lstlisting}
This effect can be handled via the local state effect, where the references track the number of balls in the urns.
\begin{lstlisting}
handler {
  $urn -> $loc,
  newurn _ k -> k (!ref (1, 1)),
  sample urn k ->
    do (red, blue) <- !get urn in
    do isRedBall <- !bernoulli (red / (red + blue)) in
    do incRed <- boolToInt isRedBall in
    do _ <- !set (a, (red + incRed, blue + (1 - incRed))) in
    k isRedBall
}
\end{lstlisting}
Here, the parameter parameter instantiation is to another parameter \lstinline|$urn -> $loc|, rather than a comcrete type,
composing the urn abstraction with the local state abstraction.
The effect operation \lstinline|!bernoulli : Double ~> Bool| is an inbuilt \cambria{} command that samples from a Bernoulli distribution.
Polya famously showed that the distribution of the urn is equivalent to a distribution given by first uniformly sampling from the interval
between 0 and 1, and then constantly sampling from a Bernoulli distribution with the resulting random variable instead.
Hence, we can bypass local state to produce a more efficient implementation of the Polya urn effect by instantiating \lstinline|$urn| as a double.
\begin{lstlisting}
handler {
  $urn -> Double,
  newurn _ k -> k (!uniform ()),
  sample a k -> k (!bernoulli a)
}
\end{lstlisting}
The client code is identical in both cases. It sees only \lstinline{$urn} and the operations \lstinline{!newurn} and \lstinline{!sample}.
Parametricity guarantees that it cannot distinguish the two implementations.

\subsection{Dynamically Threaded Concurrency} \label{sec:concurrency}

All of the previous examples can be expressed as individual instances of the same global effect.
The following example of concurrent thread management~\cite{kammar2025equational} is not such a case.
Instances interact non-trivially with one another and so cannot be expressed via multiple copies of handlers.

A concurrent program has four operations: \lstinline{!act} performs an observable action, \lstinline{!fork} spawns a child thread,
\lstinline{!wait} stalls the thread unill other threads have finished, and \lstinline{!stop} terminates the current thread.
The thread identifiers \lstinline|$tids| are abstract and so cannot be forged, compared, or inspected.

The following program constructs an N-shaped dependency between four actions.

\begin{lstlisting}
declare !fork : Unit ~> $tid + Unit.
declare !wait : $tid ~> Unit.
declare !act  : Str ~> Unit.
declare !stop : Unit ~> Unit.
case !fork () of {
  inl a1 -> case !fork () of {
    inl a2 -> !wait a2 ; case !fork () of {
      inl a3 -> !act "Action 4" ; !stop (),
      inr _  -> !wait a1 ; !act "Action 3" ; !stop ()
    },
    inr _  -> !act "Action 2" ; !stop ()
  },
  inr _ -> !act "Action 1" ; !stop ()
}
\end{lstlisting}

Only parent have access to the thread IDs of the children, and thus a child may only be waited on by its ancestors.
So a simple valid implementation of this concurrency effect would be to first run the continuation for the child thread, and then run the continuation of the parent thread.
Waiting becomes a no-op, so we can instantiate the thread IDs to the unit type because there is no need to track them.
The type \lstinline|A| is handled into \lstinline|Unit + A| to indicate if the thread returned a value or terminated.
\begin{lstlisting}
handler {
  $tid -> Unit,
  return x -> return (inr x),
  act  s k -> !print s ; k (),
  wait _ k -> k (),
  stop _ k -> return (inl ()),
  fork _ k -> k (inr ()) ; k (inl ())
}
\end{lstlisting}
Alternatively, we may want to visualise understand the dependencies between the observable actions for diagnostics.
We can implement a handler that converts the program into a DAG in graphviz format.
This is achieved by handling the concurrency effect into a simpler labelled poset algebra~\cite{kammar_handlers_2013} that is parametrized by a type of nodes.
\begin{lstlisting}
declare !get  : $loc ~> List($node).
declare !set  : $loc * List($node) ~> Unit.
declare !ref  : List($node) ~> $loc.
declare !node : List($node) * Str ~> List($node).
let node_handler = 
  do loc <- !ref [] in
  return handler {
    $tid -> List($node),
    return x -> return (inr x),
    act s k ->
      do deps <- !get loc in
      do a <- !node (deps, s) in
      !set (loc, a) ; k (),
    wait deps' k ->
      do deps <- !get loc in
      !set (loc, concat deps deps') ; k (),
    stop _ k ->
      return (inl ()),
    fork _ k ->
      do saved <- !get loc in
      do _ <- k (inr ()) in
      do child_deps <- !get loc in (
        !set (loc, saved) ;
        k (inl child_deps)
      )
  }
\end{lstlisting}
This handling scheme also shows that we may interleave handlers with multiple parameters without worry.
The local state effect is storing the sets of abstract nodes in its memory locations.
We can finally implement the node effect interface by printing the nodes and dependencies in graphviz format, using the fresh name handler from \Cref{sec:local-state}.
\begin{lstlisting}
handler {
  $node -> Unique,
  return x -> return x,
  node (deps, s) k ->
    do a <- !unique () in (
      for (fun d -> !print (hash d ++ " -> " ++ hash a)) deps ;
      !print (hash a ++ " [label=\"" ++ s ++ "\"]") ;
      k (a :: [])
    )
}
\end{lstlisting}
This example cannot be expressed using effect instances or existential types because the \lstinline{fork} operation
uses multi-shot continuations to run both child and parent threads. The thread identifiers must pass between threads while remaining abstract.
The parameter must genuinely \emph{vary}, rather than being a fixed existential witness.

%% ═══════════════════════════════════════════════════════════════
\section{The Calculus}\label{sec:calculus}
%% ═══════════════════════════════════════════════════════════════

We now introduce the calculus underpinning \cambria{}.
For clarity, we chose to present only a minimal set of language constructs that highlight the interactions of parametrized handlers.
The full language includes integers, strings, product types, sum types and more.
All of these are standard extensions orthogonal to the theory of parametrized handlers.
All metatheoretic results extend straightforwardly.

\subsection{Types and Syntax}

The syntax of the language is the following.
\[
\begin{array}{lrcl}
  \text{Value types}           & A, B   & ::= & \alpha \mid \psi \mid \tbool \mid \tfun{A}{C} \mid \thandler{C}{D} \\[4pt]
  \text{Computation types}     & C, D   & ::= & \comp{A}{\Sig} \\[4pt]
  \text{Effect signatures}     & \Sig   & ::= & \emptyset \mid \Sig, \op : A \to B \\[4pt]
  \text{Type variable context} & \Theta & ::= & \eenv \mid \Theta, \alpha \\[4pt]
  \text{Parameter context}     & \Psi   & ::= & \eenv \mid \Psi, \psi \\[4pt]
  \text{Term context}          & \G     & ::= & \eenv \mid \G,\; x : \forall \vec{\alpha} . A
\end{array}
\]
There are three contexts: $\Theta$ for type variables, $\Psi$ for parameter variables, and $\Gamma$ for term variables.
Notably, the only extension to the type system beyond a standard handler calculus is the parameter variables $\psi$.
These types can be thought of as existential type variables, because they represent an implementation detail in the effect interface.

The syntax of values and computations is the following.
\[
\begin{array}{lrcl}
  \text{Values}       & v, w & ::= & x \mid \ttrue \mid \ffalse
                                     \mid \fun x \mapsto c \mid \recmark f\;x \mapsto c \mid h \\[4pt]
  \text{Handlers}      & h    & ::= &
    \hhandler\;\{
      \ret x \mapsto c_r,\;
      \op_1\; x\; k \mapsto c_1,\; \cdots,\;
      \op_m\; x\; k \mapsto c_m
  \} \\[4pt]
  \text{Computations} & c, d & ::= & \ret v \mid \op(v;\;x.c)\mid \ddo x \leftarrow c \iin d 
                                     \mid \iif v \then c_1 \eelse c_2 \\[2pt]
                      &      &     & \mid\; v_1\;v_2 \mid \with \sigma, v \handle c \\[4pt]
  \text{Parameter subst.} & \sigma & ::= & \emptyset \mid \sigma[\psi \mapsto A]
\end{array}
\]
We write $\hhandler\;\{ \ret x \mapsto c_r,\; \op\; x\; k \mapsto c_\op \}_{\op\in\Sig}$ as shorthand for the handler syntax where
$\Sig = \{\op_1 : A_1 \to B_1, \; \cdots\;, \op_m : A_m \to B_m \}$.
The surface syntax $!\op\;v$ desugars to $\op(v;\;x.\ret x)$.
The key novelty is the \emph{parameter substitution}, often written $\sigma = \{\vec{\psi} \mapsto \vec{A}\}$, in the handle term,
which bridges the abstract types in the body to their concrete representations in the handler.

\subsection{Typing Rules}\label{sec:typing}

A type judgement $\tj \Theta \Psi A$ states that the type $A$ is well-formed in type varaible context $\Theta$ and parameter context $\Psi$.
Well-formed value types, computation types, schemes and environments are defined inductively.
\begin{mathpar}
  
  \inferrule{\alpha \in \Theta}{\tj \Theta \Psi \alpha}

  \inferrule{\psi \in \Psi}{\tj \Theta \Psi \psi}

  \inferrule{ }{\tj \Theta \Psi \tbool}

  \inferrule{ \tj \Theta \Psi A \and \tj \Theta \Psi C }{\tj \Theta \Psi \tfun{A}{C}}

  \inferrule{ \tj \Theta \Psi C \and \tj \Theta \Psi D }{\tj \Theta \Psi \thandler{C}{D}}

  \inferrule{ [\tj \Theta \Psi {A_\op} \quad \tj \Theta \Psi {B_\op} ]_{(\op : A_\op \to B_\op) \in \Sig} }{\tj \Theta \Psi \Sig}

  \inferrule{ \tj \Theta \Psi A \and \tj \Theta \Psi \Sig }{\tj \Theta \Psi {\comp{A}{\Sig}}}

  \inferrule{ \tj {\Theta, \vec{\alpha}} \Psi A }{\tj \Theta \Psi {\forall\vec\alpha.A}}

  \inferrule{ [\tj \Theta \Psi {\forall\vec\alpha.A}]_{(x:\forall \vec\alpha.A) \in \Gamma} }{\tj \Theta \Psi \Gamma}
\end{mathpar}
Value judgements are of the form $\vj \Theta \Psi \Gamma v A$ and implicitly enforce that $\tj \Theta \Psi \Gamma$ and $\tj \Theta \Psi A$.
They are defined inductively over the language syntax.
\begin{mathpar}
  \inferrule
  {(x : \forall \vec{\alpha}.B) \in \G \\ [\tj \Theta \Psi {A_\alpha}]_{\alpha\in\vec\alpha}}
  {\vj{\Theta}{\Psi}{\G}{x}{B[A_\alpha / \alpha]_{\alpha\in\vec\alpha}}}
  \; (\textsc{Var})

  \inferrule
    {b \in \{\ttrue, \ffalse\}}
    {\vj{\Theta}{\Psi}{\G}{b}{\tbool}}
    \; (\textsc{Bool})

  \inferrule
    {\cjp{\Theta}{\Psi}{\G, x : A}{c}{C}}
    {\vj{\Theta}{\Psi}{\G}{\fun x \mapsto c}{\tfun{A}{C}}}
    \; (\textsc{Fun})

  \inferrule
    {\cj{\Theta}{\Psi}{\G, f : \tfun{A}{\comp{B}{\Sig}},\, x : A}{c}{B}{\Sig}}
    {\vj{\Theta}{\Psi}{\G}{\rec f\; x \mapsto c}{\tfun{A}{\comp{B}{\Sig}}}}
    \; (\textsc{Rec})

  \inferrule{
    \cj{\Theta}{\Psi}{\G, x : A}{c_r}{B}{\Sig'}
    \\
    \Sig \setminus \Sig'' \subseteq \Sig'
    \\\\
    \left[
      \cj{\Theta}{\Psi}{\G,\, x : A_\op,\, k : \tfun{B_\op}{\comp{B}{\Sig'}}}{c_\op}{B}{\Sig'}
    \right]_{(\op:A_\op \to B_\op) \in \Sig''}
    }
    {\vj{\Theta}{\Psi}{\G}{
      \hhandler\;\{\ret x \mapsto c_r,\;
        \op\; x\; k \mapsto c_\op
      \}_{\op \in\Sig''}
    }{\thandler{\comp{A}{\Sig}}{\comp{B}{\Sig'}}}}
    \; (\textsc{Handler})
\end{mathpar}
The handled operations do not need to cover the full set of operations in the computation effect signature.
The remaining operations are forwarded, which is indicated by the $ \Sig \setminus \Sig'' \subseteq \Sig' $
premise in the \textsc{Handler} rule.

Computation judgements are of the form $\cj \Theta \Psi \Gamma c A \Sig$
and implicitly enforce that $\tj \Theta \Psi \Gamma$ and $\tj \Theta \Psi \comp{A}{\Sig}$.
In particular, all $\alpha \in \vec\alpha$ do not appear free in $\Gamma$ or $\Sig$ in the \textsc{Gen} rule.
This permits let-polymorphism without any value restriction~\cite{kammar_no_2017}.
Also, in the \textsc{HandleParam} rule, while the body $c$ is typed in the extended parameter context $\Psi, \vec{\psi}$,
the handler'soutput type $D$ is in the outer context $\Psi$, which does not contain $\vec{\psi}$, preventing the parameters from escaping.
The handler's input type is the same as the body's, but with the parameters instantiated via the parameter substitution.
When $\vec{\psi} = \emptyset$, the rule reduces to standard handler application.
\begin{mathpar}
  \inferrule
  {\vj{\Theta}{\Psi}{\G}{v}{A}}
  {\cj{\Theta}{\Psi}{\G}{\ret v}{A}{\Sig}}
    \; (\textsc{Return})

  \inferrule{
    \cj {\Theta, \vec{\alpha}} \Psi \G c A \Sig
  }{
    \cj \Theta \Psi \G c {\forall \vec{\alpha}.A} \Sig
  }
  \; (\textsc{Gen})

  \inferrule
  {\cj{\Theta}{\Psi}{\G}{c_1}{\forall \vec{\alpha}. A}{\Sig}
    \\
    \cj{\Theta}{\Psi}{\G, x : \forall\vec{\alpha}.A}{c_2}{B}{\Sig}
    }
    {\cj{\Theta}{\Psi}{\G}{\ddo x \leftarrow c_1 \iin c_2}{B}{\Sig}}
    \; (\textsc{Do})

  \inferrule
    {(\op : \arity{A_{\op}}{B_{\op}}) \in \Sig
    \\
    \vj{\Theta}{\Psi}{\G}{v}{A_{\op}}
    \\
    \cj{\Theta}{\Psi}{\G,\, x : B_{\op}}{c}{C}{\Sig}
    }
    {\cj{\Theta}{\Psi}{\G}{\op(v;\;x.c)}{C}{\Sig}}
    \; (\textsc{Op})

  \inferrule
  {\vj{\Theta}{\Psi}{\G}{v}{\tbool}
    \\
    \cj{\Theta}{\Psi}{\G}{c_1}{A}{\Sig}
    \\
    \cj{\Theta}{\Psi}{\G}{c_2}{A}{\Sig}
    }
    {\cj{\Theta}{\Psi}{\G}{\iif v \then c_1 \eelse c_2}{A}{\Sig}}
    \; (\textsc{If})

  \inferrule
  {\vj{\Theta}{\Psi}{\G}{v_1}{\tfun{A}{C}}
    \\
    \vj{\Theta}{\Psi}{\G}{v_2}{A}
    }
    {\cjp{\Theta}{\Psi}{\G}{v_1\; v_2}{C}}
    \; (\textsc{App})

  \inferrule{
    \vj{\Theta}{\Psi}{\G}{v}{\thandler{C[\sigma]}{D}}
    \\
    \cjp{\Theta}{\Psi, \dom(\sigma)}{\G}{c}{C}
    \\
    \left[\tj \Theta \Psi A\right]_{A\in\cod(\sigma)}
    }{
      \cjp{\Theta}{\Psi}{\G}{\with v, \sigma \handle c}{D}
    }
    \; (\textsc{HandleParam})
\end{mathpar}
Parameters are never generalised and must remain rigid within the body of a handler so that the handler can instantiate them consistently.
Otherwise the abstraction boundary would break.

\subsection{Operational Semantics}

Evaluation contexts identify where the next reduction step occurs:
\[
  E \;::=\; \square
    \mid \ddo x \leftarrow E \iin c
    \mid \with v \handle E
\]
Let $h = \hhandler\;\{\ret x \mapsto c_r,\;\op\;x\;k \mapsto c_\op\}_{\op\in\Sig}$.
Single-step reduction $c \leadsto c'$ is defined below, closed under the congruence rule: if $c \leadsto c'$ then $E[c] \leadsto E[c']$.
\[
\begin{array}{lcll}
  (\fun x \mapsto c)\;v
    & \leadsto & c[v/x]
    & (\beta\text{-fun}) \\
  (\rec f\;x \mapsto c)\;v
    & \leadsto & c\bigl[v/x,\;(\rec f\;x \mapsto c)/f\bigr]
    & (\beta\text{-rec}) \\
  \ddo x \leftarrow \ret v \iin c
    & \leadsto & c[v/x]
    & (\beta\text{-do-ret}) \\
  \ddo x \leftarrow \op(v; y.c) \iin c'
    & \leadsto & \op(v; y.\ddo x \leftarrow c \iin c')
    & (\beta\text{-do-op}) \\
  \iif \mathsf{true} \then c_1 \eelse c_2
    & \leadsto & c_1
    & (\beta\text{-if}_T) \\
  \iif \mathsf{false} \then c_1 \eelse c_2
    & \leadsto & c_2
    & (\beta\text{-if}_F) \\
  \with h, \sigma \handle \ret v
    & \leadsto & c_r[v/x]
    & (\beta\text{-ret}) \\
    \with h,\sigma \handle \op(v;\;y.c)
    & \leadsto & c_\op\bigl[v/x,\;(\fun y \mapsto \with h, \emptyset \handle c)/k\bigr]
    & (\beta\text{-op}) \\
  \with h,\sigma \handle \op(v;\;y.c)
    & \leadsto & \op(v;\;y.\;\with h, \emptyset \handle c)
    & (\beta\text{-fwd})
\end{array}
\]
The ($\beta$-op) rule applies when $\op\in\ops(h)$, while the ($\beta$-fwd) rule applies when $\op \notin\ops(h)$.
The continuation supplied to the operation clause in ($\beta$-op) reinstates the handler, giving \emph{deep} handling.
In the reinstated continuations, we replace the parameter substitution $\sigma$ with $\emptyset$ because the parameters are only needed for
type checking and do not have an effect on the reductions, as formalised in \Cref{lem:erasure}.

The parameter substitution does not appear in any reduction result, thus the reduction sequence is independent of these substitutions.
This is the formal sense in which parameter variables are erased at runtime. There is no dynamic dispatch, coercions, or type-directed reduction.
The parameter annotations are only used for type checking.

\begin{proposition}[Parameter Erasure]\label{lem:erasure}
  Define a parameter erasure map $\lfloor {-} \rfloor$ as
  \[
    \lfloor \with h, \sigma \handle c \rfloor = \with \lfloor h \rfloor, \emptyset \handle \lfloor c \rfloor
  \]
  and structural identity elsewhere. Then $c \leadsto c' \iff \lfloor c \rfloor \leadsto \lfloor c' \rfloor$.
\end{proposition}

\subsection{Type Safety}

For the given effect system, we have the following type safety result.

\begin{theorem}[Safety]\label{thm:safety}
  If $\cj{\eenv}{\eenv}{\eenv}{c}{A}{\Sig}$, then either:
  \begin{itemize}
    \item $c = \ret v$ for some $\vj{\eenv}{\eenv}{\eenv}{v}{A}$
    \item $c = \op(v;\; y.c')$ for some $\op : A_\op \to B_\op \in \Sig$, $\vj{\eenv}{\eenv}{\eenv}{v}{A_\op}$,
      and $\cj{\eenv}{\eenv}{y:B_\op}{c}{A}{\Sig}$.
    \item $c \leadsto c'$ for some $\cj{\eenv}{\eenv}{\eenv}{c'}{A}{\Sig}$
  \end{itemize}
\end{theorem}

\begin{proof}
  We first prove type substitution, parameter substitution and variable substitution lemmas by structural induction:
  \begin{itemize}
    \item If $\cj{\Theta,\alpha}{\Psi}{\Gamma}{c}{A}{\Sig}$ and $\Theta;\Psi \vdash B$, then $\cjp{\Theta}{\Psi}{\Gamma[B / \alpha]}{c[B/ \psi]}{(\comp{A}{\Sig})[B/ \alpha]}$.
    \item If $\cj{\Theta}{\Psi,\psi}{\Gamma}{c}{A}{\Sig}$ and $\Theta;\Psi \vdash B$, then $\cjp{\Theta}{\Psi}{\Gamma[B / \psi]}{c[B/ \psi]}{(\comp{A}{\Sig})[B/ \psi]}$.
    \item If $\cj{\Theta}{\Psi}{\Gamma, x:\forall \vec{\alpha}.B}{c}{A}{\Sig}$ and $\vj{\Theta,\vec{\alpha}}{\Psi}{\Gamma}{v}{A}$, then $\cjp{\Theta}{\Psi}{\Gamma}{c[v / x]}{\comp{A}{\Sig}}$.
  \end{itemize}
  Progress is then shown by induction over the computation derivations, while preservation is shown by case analysis on the operational steps.
\end{proof}

\subsection{Contextual Equivalence}

Two programs in \cambria{} are contextually equivalent when there are no contexts that they can be place in that distinguishes their behaviour.

\begin{definition}
  A \emph{computation context} is a computation with a hole in it.
  \[
  \begin{array}{lrcl}
    \text{Computation contexts} & \ctx & ::= & \square \mid \op(v;\;x.\ctx) \mid \ddo x \leftarrow \ctx \iin c
                                       \mid \ddo x \leftarrow c \iin \ctx
                                       \\[2pt]
                        &      &     & \mid \iif v \then \ctx \eelse c \mid \iif v \then c \eelse \ctx \mid\; \with \sigma, v \handle \ctx
  \end{array}
  \]
  Given a well-typed computation $\cj \Theta \Psi \Gamma c A \Sig$,
  a \emph{closing context} for $c$ is a computation context $\ctx$ such that $\cj \eenv \eenv \eenv {\ctx[c]} \tbool \emptyset$.
  Given two well-typed computations $\cj \Theta \Psi \Gamma {c_1, c_2} A \Sig$, they are \emph{contextually equivalent}, written $c_1 \approx c_2$,
  if for all closing contexts $\ctx$,
  \[
    \ctx[c_1] \leadsto^* \ret b \iff \ctx[c_2] \leadsto^* \ret b
  \]
  where $b \in \{\ttrue, \ffalse\}$.

\end{definition}

%% ═══════════════════════════════════════════════════════════════
\section{Parametricity}\label{sec:parametricity}
%% ═══════════════════════════════════════════════════════════════

In this section we prove parametricity for the parameters and type variables.
Code in contexts $\Theta$ and $\Psi$ cannot inspect the concrete representation of the variables.
We prove this result using a step-indexed binary logical relation.
The key insight with this proof is that parametrized handers can be decomposed into plain handlers and existential types.
With this decomposition, the logical relation becomes a standard combination of handlers, universal types, existential types and recursion,
justifying the canonicity of parametrized handlers.
The parametricity result is shown to carry over to the surface language via an elaboration proof.
Viewing the parameters as existential types means that we do not need to track the dynamic resources as an effect creates them, we only posit one exists at the end of the computation.
This means we avoid the complex machinery of Kripke worlds or biorthogonality, making the language more adaptable to further extensions like
recursive types, linear types and universally quantified effects.

\subsection{The Core Language}

Core to the parametricity proof is the insight that parametrized handlers decompose into plain handlers and existential types.
In particular, for a single parameter substitution, $\sigma = \{\psi \mapsto A\}$,
\begin{align*}
  \with v, \{\psi \mapsto A\} \handle c \quad = &\quad \unpack \pack(\psi, A, v) \as (\psi, x) \iin (\with x \handle c).
\end{align*}
This decomposition is formalised in \Cref{sec:elaboration}. Before proceeding to that, we introduce the syntax of the core language.
We update the handle syntax and extend the value types, values and computations with the following constructs in replace of the parametrized handle term.
\[
\begin{array}{lrcl}
  \text{Value types}  & A, B & ::= & \cdots \mid \exists\psi.A \\[4pt]
  \text{Values}       & v, w & ::= & \cdots \mid \pack(\psi, A, v) \\[4pt]
  \text{Computations} & c, d & ::= & \cdots \mid \cancel{\with v, \sigma \handle c} \mid \with v \handle c \mid \unpack v \as (\psi, x) \iin c
\end{array}
\]
Parameter substitutions $\sigma$ are no longer part of the syntax in the core language.
We have a new typing derivation for existential types, and in place of the single \textsc{HandleParam} rule,
we have a plain handle rule, as well as existential introduction and elimination rules.
\begin{mathpar}
  \inferrule{ \tj \Theta {\Psi, \psi} A }{ \tj \Theta \Psi {\exists\psi.A}}

  \inferrule{
    \vj{\Theta}{\Psi}{\G}{v}{\exists \psi.A}
    \\
    \vj{\Theta}{\Psi,\psi}{\G, x:A}{c}{C}
  }{
    \cjp{\Theta}{\Psi}{\G}{\unpack v \as (\psi, x) \iin c}{C}
  }
  \; (\textsc{Unpack})

  \inferrule{
    \vj{\Theta}{\Psi}{\G}{v}{\thandler{C}{D}}
    \\
    \cjp{\Theta}{\Psi}{\G}{c}{C}
  }{
    \cjp{\Theta}{\Psi}{\G}{\with v \handle c}{D}
  }
  \; (\textsc{Handle})

  \inferrule{
    \tj{\Theta}{\Psi}{A}
    \\
    \vj{\Theta}{\Psi}{\G}{v}{B[A / \psi]}
  }{
    \vj{\Theta}{\Psi}{\G}{\pack(\psi, A, v)}{\exists \psi.B}
  }
  \; (\textsc{Pack})
\end{mathpar}
Note that $\psi \notin \fpv(C, \G)$ in the \textsc{Unpack} rule because $\tj{\Theta}{\Psi}{\G}$ and $\tj{\Theta}{\Psi}{C}$ implicitly in the conclusion.

The operational semantics replaces parametrized handler reduction with classic reduction and adds a new reduction rule for existential terms.
\[
\begin{array}{lcll}
  \with h \handle \ret v
    & \leadsto_c & c_r[v/x]
    & (\beta\text{-ret}) \\
  \with h \handle \op_i(v;\;y.c)
    & \leadsto_c & c_i\bigl[v/x,\;(\fun y \mapsto \with h \handle c)/k\bigr]
    & (\beta\text{-op}) \\
  \with h \handle \op(v;\;y.c)
    & \leadsto_c & \op(v;\;y.\;\with h \handle c)
    & (\beta\text{-fwd}) \\
  \unpack \pack(\psi, A, v) \as (\psi, x) \iin c
    & \leadsto_a & c[v / x]
    & (\beta\text{-ext})
\end{array}
\]
To make it clear which operational semantics we are using, we denote reduction steps in the core language
by $\leadsto_c$ if it is a core reduction step, and by $\leadsto_a$ if it is an administrative reduction step.
The only administrative reduction step is ($\beta$-ext), which of no computational consequence because it
instantiates existential types, which have no runtime impact.
In \cambria{}, this is captured by the single parametrized handle reduction, as we formalise in \Cref{prop:op-corresp}.

\subsection{Step-Indexed Logical Relation and Fundamental Theorem}

In this section, we build the logical relation to prove parametricity of \cambria{}. The definitions of the logical relation are standard.
The step-indexing integrates nicely with the handler relation by making the self-reference in the typing derivation well-founded,
rather than having to revert to inducting on the size of the computation tree. While we only introduce the relation on the core syntax,
it extends readily to product types, sum types, list types, map types, and all of the current features in the current version of \cambria{}.
The ease of integration primarily comes from the decomposition of parametrized handlers into plain handlers and existentials, making each induction case canonical.

We build up a relational model of contextual equivalence via a step-indexed logical relation.
The relation is defined mutually inductively over values, computation, schemes and environments.
The step-index tracks whether two terms cannot be discerned within that number of steps.
Thus, the relation should only grow with the size of the index.

\begin{definition}
  An \emph{admissible relation} is a step-indexed relation on closed value terms
  $R \subseteq \mathsf{Val}_0 \times \mathsf{Val}_0 \times \mathbb{N}$, an element of which we write $(v_1, v_2)_n$,
  that is downward-closed. If $(v_1, v_2)_n \in R$ and $n' \leq n$, then $(v_1, v_2)_{n'} \in R$.
  We write $\Rel$ for the set of all admissible relations.
\end{definition}

Given a type context $\Theta$ and parameter context $\Psi$,
a \emph{relational interpretation} $\delta : (\Theta \cup \Psi) \to \Rel$ assigns an admissible relation to each type variable and parameter.
The logical relation assigns to each typing judgement $\tj \Theta \Psi X$, a mapping from relational interpretations to an admissible relation.
On value types, this is written as $\VV\llbracket X \rrbracket_\delta$.
We also qualify with $\CC$, $\SSS$, and $\GG$ to make it clear whether we are interpreting values, computations, schemes or environments.
It is defined inductively over the typing judgements, and a step index $n\in\mathbb{N}$.
\begin{align*}
  \VV\llbracket \alpha \rrbracket_\delta &= \delta(\alpha) \\
  \VV\llbracket \psi \rrbracket_\delta &= \delta(\psi) \\
  \VV\llbracket \tbool \rrbracket_\delta &= \{(b, b)_n \mid b \in \{\ttrue, \ffalse\}, n \in \mathbb{N}\} \\
  \VV\llbracket \tfun{A}{C} \rrbracket_\delta &= \{(v_1, v_2)_n \mid \forall j \leq n.\; \forall (w_1, w_2)_j \in \VV\llbracket A \rrbracket_\delta.\; (v_1\;w_1, v_2\;w_2)_j \in \CC\llbracket C \rrbracket_\delta\} \\
  \VV\llbracket \thandler{C}{D} \rrbracket_\delta &= \{(v_1, v_2)_n \mid \forall j \leq n.\; \forall (c_1, c_2)_j \in \CC\llbracket C \rrbracket_\delta. \\
  &\qquad\qquad\qquad (\with v_1 \handle c_1, \with v_2 \handle c_2)_j \in \CC\llbracket D \rrbracket_\delta\} \\
  \VV\llbracket \exists\psi.A \rrbracket_\delta &= \{(\pack(\psi, B_1, v_1), \pack(\psi, B_2, v_2))_n \mid \exists R \in \Rel.\; (v_1, v_2)_n \in \VV\llbracket A \rrbracket_{\delta[\psi \mapsto R]}\} \\
  \SSS\llbracket \forall\vec\alpha.A \rrbracket_\delta &= \{(v_1, v_2)_n \mid \forall \vec{R} \in \Rel^{|\vec\alpha|}.\; (v_1, v_2)_n \in \VV\llbracket A \rrbracket_{\delta[\vec\alpha \mapsto \vec{R}]}\} \\
  \GG\llbracket \G \rrbracket_\delta &= \{(\gamma_1, \gamma_2)_n \mid \forall(x:\forall\vec\alpha.A)\in\Gamma. \; (\gamma_1(x), \gamma_2(x))_n \in\SSS\llbracket \forall\vec\alpha.A \rrbracket_\delta \}
\end{align*}
For two closed computations, they are related at step index $n$ if they remain observationally equivalent after $n$ reduction steps.
That is, $(c_1,c_2)_n \in \CC\llbracket \comp{A}{\Sig} \rrbracket_\delta$ if for all $j < n$:
\begin{itemize}
  \item If $c_1 \leadsto^j \ret v_1$, then there is $v_2$ such that $c_2 \leadsto^* \ret v_2$ and $(v_1, v_2)_{n-j} \in \VV\llbracket A \rrbracket_\delta$
  \item If $c_1 \leadsto^j \op(v_1; y.d_1)$ for $(\op : \arity{A_\op}{B_\op}) \in \Sig$, then there is $v_2$ and $d_2$ such that $c_2 \leadsto^* \op(v_2; y.d_2)$ and:
  \begin{itemize}
    \item $(v_1, v_2)_{n-j} \in \VV\llbracket A_\op \rrbracket_\delta$
    \item $\forall k < n{-}j.\; \forall (w_1, w_2)_k \in \VV\llbracket B_\op \rrbracket_\delta.\; (d_1[w_1/y], d_2[w_2/y])_k \in \CC\llbracket \comp{A}{\Sig} \rrbracket_\delta$
  \end{itemize}
\end{itemize}
The step index tracks reduction steps on the left.
The self-reference of $\CC\llbracket \comp{A}{\Sig} \rrbracket_\delta$ is well-founded because the index strictly reduces.

\begin{theorem}[Parametricity]\label{thm:parametricity}
  If $\cjp{\Theta}{\Psi}{\G}{c}{C}$, then for all $\delta : (\Theta \cup \Psi) \to \Rel$, $n\in\mathbb{N}$,
  and $(\gamma_1, \gamma_2)_n \in \GG\llbracket \G \rrbracket_\delta$,
  then $(c[\gamma_1], c[\gamma_2])_n \in \CC\llbracket C \rrbracket_\delta$.
\end{theorem}

\begin{proof}
By induction on the typing derivation. 

\emph{Rec:} By inner induction on $n$.
For $n \geq 1$, the inner induction hypothesis gives the recursive function related at $n{-}1$.
The fundamental theorem (at $n{-}1$) then gives the function bodies related at $n{-}1$,
and the anti-reduction lemma lifts this relation to $n$.

\emph{Handler:} By inner induction on $n$.
Two related computations will the return case applies the fundamental theorem to $c_r$;
the operation case constructs the deep continuation $\fun y \mapsto \with h \handle c'$,
which is related at $\VV_{m+1}\llbracket \tfun{B_\op}{\comp{B}{\Sig'}} \rrbracket$ by the inner induction hypothesis at $j' \leq m$.

\emph{Handle:} Immediate from the handler relation at $i = n$.

\emph{Do:} Via a separate Bind Lemma, proved by strong induction on $n$:
if $(c_1, c_2)_n \in \CC\llbracket \comp{A}{\Sig} \rrbracket$ and for all $i \leq n$ and $(v_1, v_2)_n \in \VV\llbracket A \rrbracket$
we have $(d_1[v_1/x], d_2[v_2/x])_i \in \CC\llbracket \comp{B}{\Sig} \rrbracket$,
then $(\ddo x \leftarrow c_1 \iin d_1, \ddo x \leftarrow c_2 \iin d_2)_n \in \CC\llbracket \comp{B}{\Sig} \rrbracket$.

\emph{Gen:} Since reduction is independent of the relational interpretation $\delta$, the same witnesses work for all instantiations $\vec{R}$.
\end{proof}

\todo[inline]{Expand this proof}

\subsection{Elaboration and Adequacy} \label{sec:elaboration}

\newcommand{\elab}[1]{{(#1)^\dagger}}

This section demonstrates that \cambria{} elaborates faithfully into the core language.
Hence, the fundamental theorem, \Cref{thm:parametricity}, is reflected by \cambria{},
and the abstract interfaces of the parametrized handlers are formally guaranteed.

\begin{definition}
  The \emph{elaboration translation} $\elab{{-}}$ from \cambria{}'s syntax to the core syntax is given by
  structural identity on terms, apart from the parametrized handle clause, where it is defined inductively over the size of the parameter substitution.
  \begin{align*}
    \elab{\with v, \emptyset \handle c}              &= \with v^\dagger \handle c^\dagger \\[-2pt]
    \elab{\with v, \sigma[\psi \mapsto A] \handle c} &= \unpack \pack(\psi, A, v^\dagger) \as (\psi, x) \iin \elab{\with x, \sigma \handle c}
  \end{align*}
\end{definition}

This elaboration preserves typing judgements, and is thus well-defined.

\begin{proposition}[Type Preservation]
  If $\cjp \Theta \Psi \Gamma c C$ in \cambria{}, then $\cjp \Theta \Psi \Gamma {c^\dagger} C$ in the core syntax.
\end{proposition}

\begin{proof}
  The only non-trivial case is the \textsc{Handle} elaboration. This is proved by induction on the size of the parameter substitution.
  When $\sigma=\emptyset$, this reduces to the classic handler case:
  \[
  \inferrule{
    \vj{\Theta}{\Psi}{\G}{v}{\thandler{C}{D}}
    \\
    \cjp{\Theta}{\Psi}{\G}{c}{C}
    }{
      \cjp{\Theta}{\Psi}{\G}{\with v, \emptyset \handle c}{D}
    }
  \]
  When $\sigma=\sigma'[\psi \mapsto A]$, unwinding the typing derivation gives:
  \[
    \inferrule{
      \vj{\Theta}{\Psi}{\G}{v}{\thandler{C[\sigma'[\psi\mapsto A]]}{D}}
      \\
      \cjp{\Theta}{\Psi,\dom(\sigma'), \psi}{\G}{c}{C}
      \\
      \tj \Theta \Psi A \quad \cdots
    }{
      \cjp{\Theta}{\Psi}{\G}{\with v, \sigma'[\psi \mapsto A] \handle c}{D}
    }
  \]
  As $\psi \notin \fpv(D)$, we can derive
  \begin{equation}
    \inferrule{
      \vj{\Theta}{\Psi}{\G}{v}{(\thandler{C[\sigma']}{D})[A / \psi]}
    }{
      \vj{\Theta}{\Psi}{\G}{\pack(\psi, A, v)}{\exists\psi.(\thandler{C[\sigma']}{D})}
    }\text, \label{eq:pack}
  \end{equation}
  and weakening $\cjp{\Theta}{\Psi,\dom(\sigma'), \psi}{\G}{c}{C}$ gives,
  \begin{equation}
    \cjp \Theta {\Psi, \psi} {\Gamma, x : \thandler{C[\sigma']}{D}} {\with x, \sigma' \handle c} D . \label{eq:freepsi}
  \end{equation}
  Then, combining \eqref{eq:pack} and \eqref{eq:freepsi} we can derive
  \[
    \cjp{\Theta}{\Psi}{\G}{\unpack \pack(\psi, A, v) \as (\psi, x) \iin \elab{\with x, \sigma' \handle c}}{D}
  \]
  where we invoked the IH on $\elab{\with x, \sigma' \handle c}$.
\end{proof}

This elaboration is sound in the following sense.
A term in \cambria{} reduces if and only if its elaboration reduces in the core language.
And this reduction in the core language must be of the form of any number of administrative reductions followed by a single core reduction.
That is, reduction steps in \cambria{} are one-to-one with core reductions,
along with some plumbing to handle the existential elaborations, which are handled in a single reduction step in \cambria{}.

\begin{proposition}[Operational Correspondence] \label{prop:op-corresp}
  For a closed term $\cj \eenv \eenv \eenv c A \Sig$, $c \leadsto c'$ if and only if $c^\dagger \leadsto_a^* \ \leadsto_c {c'}^\dagger$.
\end{proposition}

\begin{proof}
  The proof is given by case analysis on the reductions in \cambria{}.
  The only non-trivial cases are the parametrized handler reductions,
  because the elaboration is structural identity on all other terms.
  When the parameter substitution is empty,
  \begin{equation}
    \with h, \emptyset \handle d \leadsto c' \iff (\with h, \emptyset \handle d)^\dagger \leadsto_c c'^\dagger \label{eq:emptystep}
  \end{equation}
  because $(\with h, \emptyset \handle d)^\dagger = (\with h^\dagger \handle d^\dagger)$ and the elaboration is stable under substitution.
  If the parameter substitution is not empty, we show by induction that
  \[
    (\with h, \sigma \handle c)^\dagger \leadsto^*_a (\with h, \emptyset \handle c)^\dagger.
  \]
  For $\sigma = \emptyset$, this is immediate after zero reductions.
  For $\sigma = \sigma'[\psi \mapsto A]$,
  \begin{align*}
    (\with h, \sigma'[\psi \mapsto A] \handle d)^\dagger
    &= \unpack \pack(\psi, A, h) \as (\psi, x) \iin (\with x, \sigma' \handle d)^\dagger \\
    &\leadsto_a (\with h, \sigma' \handle d)^\dagger \\
    &\leadsto_a^* (\with h, \emptyset \handle d)^\dagger \tag{IH}
  \end{align*}
  As $(\with h, \sigma \handle d)$ reduces identically to $(\with h, \emptyset \handle d)$, the result follows from \eqref{eq:emptystep}.
\end{proof}

Finally, we may lift the parametricity result (\Cref{thm:parametricity}) from the core language to \cambria{}.

\begin{theorem}[Adequacy]
  Let $\cjp \Theta \Psi \Gamma {c_1, c_2} C$. If for all relational interpretations $\delta: (\Theta \cup \Psi) \to \Rel$,
  step-indexes $n\in\mathbb{N}$, and related substitutions $(\gamma_1, \gamma_2)_n \in \GG\llbracket \Gamma \rrbracket_\delta$,
  \[
    (c_1[\gamma_1]^\dagger, c_2[\gamma_2]^\dagger)_n \in \CC\llbracket C \rrbracket_\delta,
  \]
  then $c_1 \approx c_2$.
\end{theorem}

\subsection{Applications of Parametricity}

Parametricity is a powerful result with many consequences.
It can be used to derive many properties of the language including
representation independence, prevention of effect encapsulation, commutativity, and garbage collection results.
This section outlines the key corollaries of \Cref{thm:parametricity}.

\subsubsection{Representation Independence for Parameterized Handlers}

\begin{theorem}[Representation Independence]
  Let $\tj \Theta \Psi \sigma_1, \sigma_2$ be two well-typed assignment of types to parameters $\vec\psi$.
  Let $\{\vec\psi \mapsto \vec R\}$ be an assignment of each $\psi$ to an admissible relation.
  Then, if $(h_1, h_2)_n \in \VV\llbracket \thandler{C}{D} \rrbracket_{\delta[\vec\psi \mapsto \vec R]}$ for all $n \in \mathbb{N}$,
  then $(\with h_1, \sigma_1 \handle c) \approx (\with h_2, \sigma_2 \handle c)$.
\end{theorem}

\subsubsection{Garbage Collection for Parameterized Effects}

\subsubsection{Scope Safety of Parameters}

\subsubsection{Handler Commutativity for Independent Parameters}

%% ═══════════════════════════════════════════════════════════════
\section{Type Inference}\label{sec:inference}
%% ═══════════════════════════════════════════════════════════════

\cambria{}'s type inference algorithm extends Hindley--Milner~\cite{damas_principal_1982}
requiring explicit annotations only for operations whose signatures mention instantiated parameters.
Two completeness results relate the algorithm to the declarative system of \Cref{sec:calculus}.

\begin{lemma}[Ground Completeness] \label{thm:ground-complete}
  Suppose every handle expression in $t$ has $\sigma = \emptyset$.
  If $\cjp{\Theta}{\Psi}{\G}{t}{C}$ is derivable declaratively, the algorithm infers
  a type $C'$ of which $C$ is an instance.
\end{lemma}

\begin{proof}
  Any derivation in this fragment is identical to a derivation in a plain algebraic effects system,
  for which completeness has been established~\cite{damas_principal_1982, kammar_no_2017}.
\end{proof}

The second result covers the full parameterized system.
An \emph{annotated term} is one in which every handle expression
$\with h, \sigma \handle c$ carries an explicit type annotation on the body:
$\with h, \sigma \handle (c : C)$.

\begin{theorem}[Parameterized Completeness]\label{thm:param-complete}
  Let $c$ be an annotated term and let $\G$ contain principal types for all variables.
  If $\cjp{\Theta}{\Psi}{\G}{c}{C}$ is derivable in the declarative system, then the
  algorithm infers a principal type $C'$ of which $C$ is an instance.
\end{theorem}

\begin{proof}
  By structural induction on the typing derivation. The main cases are:
  \begin{itemize}
    \item \emph{\textsc{Do}.}
          By \emph{gen admissibility}~\cite{damas_principal_1982}, any declarative derivation
          can be rearranged so that \textsc{Gen} appears only immediately below
          \textsc{Do}. This extends to our system because effect parameters are never generalized.

    \item \emph{\textsc{Handle}.}
          The algorithm infers the handler type
          $C[\sigma] \Rightarrow D$ by the IH. The body is checked against the
          explicitly annotated type $C$ in context $\Psi, \dom(\sigma)$.
          A single substitution pass followed by equality checking compares the
          substituted body type with the handlers inferred input type.
          The result type $D$ is principal because the handler's type is principal
          by the IH, and the body's type is fixed by the annotation.
  \end{itemize}
  The remaining cases are standard.
\end{proof}

The annotation in \Cref{thm:param-complete} specifies the full type $C$ of the body.
The following theorem strengthens this to a minimal annotation consisting only of the
operation arities that mention $\dom(\sigma)$.

\begin{definition} \label{def:relevant}
  Given a parameter substitution $\sigma$, the
  \emph{$\sigma$-relevant operations} of the handled signature $\Sig$ is
  \[
    \rel_\sigma(\Sig)
    \;=\;
    \bigl\{\,\op : A_\op \to B_\op \in \Sig \mid
      \dom(\sigma) \cap (\fpv(A_\op) \cup \fpv(B_\op)) \neq \emptyset
    \,\bigr\}.
  \]
  A \emph{minimally annotated term} is one in which each handle expression carries an annotation listing only
  the $\sigma$-relevant operations, $\with h, \sigma \handle (c : \rel_\sigma(\Sig))$.
  The return type and all other operations are omitted.
\end{definition}

\begin{corollary}[Minimal Annotation Completeness]\label{thm:minimal-complete}
  Let $t$ be a minimally annotated term and let $\G$ contain principal types for all
  variables.
  If $\cjp{\Theta}{\Psi}{\G}{c}{C}$ is derivable declaratively, then the algorithm
  infers a principal type $C'$ of which $C$ is an instance.
\end{corollary}

\begin{proof}
  We extend the proof of \Cref{thm:param-complete}, strengthening the \textsc{Handle} case.
  Partition the body's effect signature into three groups:

  \begin{enumerate}
    \item \emph{$\sigma$-relevant ops.}
          The annotation seeds each $\psi \in \dom(\sigma)$ as a rigid constant
          into $\Psi$, and the argument proceeds as in \Cref{thm:param-complete}.

    \item \emph{$\sigma$-free ops.}
          For any such $\op : A_\op \leadsto B_\op$ we have
          $\dom(\sigma) \cap (\fpv(A_\op) \cup \fpv(B_\op)) = \emptyset$,
          so $\sigma(A_\op) = A_\op$ and $\sigma(B_\op) = B_\op$.
          Typing subterms against the handler's concrete arity already gives the correct
          abstract type; no annotation is needed.
          Moreover, since $\psi$ does not appear in $A_\op$ or $B_\op$, every unification
          constraint arising from $\op$ is $\sigma$-free, so the rigid $\psi$
          established by group~(1) is not disturbed.

    \item \emph{Forwarded ops.}
          Forwarded operations flow from the body's signature into $D$ unchanged, so
          their arities are $\sigma$-free.
          The argument is then identical to group~(2).
  \end{enumerate}
  \emph{Return type.}
  With each $\psi\in\dom(\sigma)$ rigid in $\Psi$, the unification variable the algorithm introduces
  for the return type is constrained only by value-level subterms of $c$.
  Since $\psi\in\dom(\sigma)$ act as a ground constants to the unifier, the resulting value type is
  principal without any annotation on the return position.
\end{proof}

The annotation burden is minimal; only operations whose types mention $\psi$ need declaration.
All other types are inferred.
Moreover, the annotation $\rel_\sigma(\Sig)$ is a term-independent sufficient condition.
Further operations in $\rel_\sigma(\Sig)$ may be omitted when $\psi$ propagates to their arity via term-level value flow.
For example, annotating only $\mathsf{get} : \unit \leadsto \psi$ in the body
$\ddo x \leftarrow {!}\mathsf{get}\; () \iin {!}\mathsf{set}\;x$
permits the algorithm to infer $\mathsf{set} : \psi \leadsto \unit$.

For this reason, Cambria allows individual effect operations to be typed with the syntax,
\begin{lstlisting}
effect op : A ~> B. c
\end{lstlisting}
but enforces no constraints on which operations need to be annotated, leaving the programmer to manage the interface.


%% ═══════════════════════════════════════════════════════════════
\section{Related Work}\label{sec:related}
%% ═══════════════════════════════════════════════════════════════

\paragraph{Algebraic effects and handlers.}
Algebraic effects were introduced by Plotkin and Power~\cite{plotkin_algebraic_2003} and extended with handlers by Plotkin and Pretnar~\cite{plotkin_handling_2009}. Many languages and calculi incorporate handlers, including Eff~\cite{bauer_programming_2015}, Koka~\cite{leijen_koka_2014}, Frank~\cite{lindley_do_2017}, Links~\cite{hillerström_continuation_2017}, and OCaml~5. \cambria{} is closest in spirit to the fine-grained call-by-value calculus of Kammar et al.~\cite{kammar_handlers_2013}, whose type safety results we extend to the parameterized setting.

\paragraph{Parameterized effects and the problem of instances.}
The need for parameterized effects --- and the inadequacy of simple effect signatures for modelling dynamically created resources --- was identified early in the algebraic effects literature. Biernacki et al.~\cite{biernacki2018handle} address this via existential types in a simply-typed calculus with biorthogonality-based logical relations, but without recursion or universal polymorphism.

\paragraph{Logical relations for effects.}
Biernacki et al.~\cite{biernacki2018handle} establish a logical relation for handlers using biorthogonality, which defines the computation relation indirectly through evaluation contexts. This gives compositionality properties for free but couples the relation to the full language syntax. Our direct, step-indexed approach is more elementary and modular: each type contributes an independent clause, and adding new features requires only local additions. Step-indexed logical relations originate with Appel and McAllester~\cite{appel_indexed_2001} and were developed for parametric polymorphism by Ahmed~\cite{ahmed_semantics_2006}. To our knowledge, our logical relation is the first to combine step-indexing with algebraic effect handlers.

\paragraph{Step-indexing and recursion.}
Step-indexed logical relations were introduced precisely to handle general recursion without requiring domain-theoretic machinery. Ahmed~\cite{ahmed_semantics_2006} establishes parametricity for System~F with recursive types. We adapt the step-indexing technique to a setting with effects and handlers, where the computation relation's self-reference through operation continuations requires careful accounting: each operation observation consumes one unit of index, ensuring well-foundedness.

\paragraph{Concurrent effects.}
Kammar et al.~\cite{kammar2025equational} develop an equational theory for dynamic threads as an algebraic effect. Their formalization uses concrete thread identifiers, without type abstraction. \cambria{}'s parameter variables make thread identifiers abstract, giving parametricity guarantees for concurrent programs that no prior system provides.

%% ═══════════════════════════════════════════════════════════════
\section{Conclusion}\label{sec:conclusion}
%% ═══════════════════════════════════════════════════════════════

We have presented \cambria{}, a calculus that extends algebraic effect handlers with parameter variables to provide type abstraction for parameterized effects. Parameter variables are first-class in the type system but erased at runtime, requiring no coercions or type-directed reduction. A step-indexed logical relation establishes parametricity: code polymorphic in a parameter variable cannot inspect its concrete representation.

The design leads to a modular metatheory. The value relation, computation relation, and fundamental theorem decompose into independent pieces, each corresponding to a single type constructor or typing rule. This modularity reflects the orthogonality of parameter variables and effect handling --- and suggests that the abstraction boundary is in the right place.

We have demonstrated \cambria{}'s expressiveness with examples spanning local state, probabilistic programming, and concurrent thread management. The concurrency example is particularly significant: it exercises a parameterized effect that cannot be reduced to effect instances, falling outside the scope of prior work on handler parametricity.

\paragraph{Future work.} Natural extensions include effect polymorphism (abstracting over effect signatures), recursive types, and a denotational semantics establishing the categorical structure of parameterized handlers. We are also interested in mechanizing the parametricity proof and exploring the implications of free theorems for handler-based program transformations.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
